M.Sc. Proposal

### 2010-09-26 notes/proposal.tex
Irad Review

\title{M.Sc. Proposal}
\author{Roman Zeyde}
\begin{document}
\maketitle
\section{Introduction}
Electrokinetic theory describes the dynamics of charged particles in ionic fluids.
When a particle acquires surface charge, a layer of ions of opposite charge is attracted to the surface via electric forces, creating a double-layer structure around the particle (see Figure \ref{fig:EDL}).
This structure, called ``Debye layer'', electrically screens the surface charge, thus creating a potential difference between the particle and the outer layer of the fluid bulk.

\begin{figure}[htbp]
\begin{framed}
    \begin{center}
        \includegraphics[width=0.25\textwidth]
            {ElectricDoubleLayer.eps}
        \caption{Schematic structure of the double layer:
        (1) particle's surface, (2) Debye layer and (3) fluid bulk.
        If the the particle's surface has positive (blue) surface charge, it attracts negative (red) ions from the fluid making the Debye layer be negatively charged (as opposed to the rest of the fluid bulk, which is electrically neutral).
        Zeta potential is defined as the voltage drop on (2).}
        \label{fig:EDL}
    \end{center}
\end{framed}
\end{figure}

In cases where the layer width is much smaller than the particle's size, an analytical solution for the Debye layer dynamics can be found \cite{Yariv10a}.

Electrokinetics is described by the following physical phenomena (in dimensionless formulation):
\begin{enumerate}
\item Electrostatic potential - the fluid bulk contains no free charge:
\begin{equation} \label{eq:Laplace}
    \bnabla \cdot (C \bnabla \varPhi) = 0.
\end{equation}
\item Fluid dynamics - Stokes incompressible flow with an electrostatic force component:
\begin{equation} \label{eq:Stokes}
\left\{ \begin{array}{l}
\bLaplacian \bV - \bnabla P + \Laplacian \varPhi \bnabla \varPhi = 0 \\
\bnabla \cdot \bV = 0 \end{array} \right.
\end{equation}
\item Ion diffusion and advection (via the fluid's velocity field):
\begin{equation} \label{eq:Nernst}
\Laplacian C - \alpha \bV \cdot \bnabla C = 0
\end{equation}
\end{enumerate}
The boundary conditions correspond to the specific problem at hand and defined by particle's geometry, chemical characteristics and fluid's dynamics.

The partial differential equations above are coupled and non-linear. In general, they have no analytic solution. Moreover, any numerical solver must handle the scale disparity, caused by the Debye layer width being much smaller than the particle's size.
A closed-form approximate solution has been developed for spherical ion-exchanging particles and small electric field \cite{Yariv10b} but it is hard to extend the analytic solution to more general systems.

\section{Research goals}
Our goal is to implement an accurate and fast iterative numerical solver for electrokinetic problems, which could be successfully applied for systems that have no closed-form solutions known (e.g. asymmetric particles or large electric fields).

The solver will be verified against known closed-form solutions for each of the sub-problems, as well as the full coupled problem, analyzing the discrepancies between the theoretical and numerical results.

Moreover, the convergence rate of the solver will be optimized using numerical acceleration methods in order to achieve high accuracy in reasonable execution time.

The solver will provide a numerical model for interesting electrokinetical phenomena which have no closed-form analytic solutions yet, such as large field dynamics and electrophoretic\footnote{Electrophoresis is the physical phenomenon of particles motion in an ion-containing fluid under the influence of an electric field.} autonomous micro-swimmers \cite{Paxton04, Howse07}.

\section{Achieved results}
The numerical solver is being implemented as a MATLAB package, which provides the needed tools for numerical solution of an electrophoresis problem.

The solver applies iterative relaxations to solve the coupled system, by
iterating the equations and relaxing each one of them
for the corresponding variable:
\begin{itemize}
  \item Laplace equation \ref{eq:Laplace} is used to update $\varPhi$.
  \item Stokes equation \ref{eq:Stokes} is used to update $\bV$ and $P$.
  \item Nernst-Planck equation \ref{eq:Nernst} is used to update for $C$.
\end{itemize}
While solving a specific equation for a specific variable, all other variables maintain their values from previous iteration, resembling ``Gauss-Seidel'' method for linear systems.

The problem's configuration is taken from \cite{Yariv10b} and consists of a spherical ion-exchanging particle, surrounded by infinite Stokes fluid.
Spherical coordinates $(R,\theta,\phi)$ system is chosen to make the problem axisymmetric, therefore independent of $\phi$.
The electrokinetic problem is written as two-dimensional partial differential equation system for $R \in [1,\infty)$ and $\theta \in [0, \pi]$.
Natural grid choices are a logarithmic grid for $R$ and an uniform grid for $\theta$.

The following linear operators are implemented in spherical coordinates on the grid described above: Scalar Laplacian (for $\varPhi$ and $C$), Vector Laplacian (for $\bV$), Gradient (for $\Phi, C$ and $P$) and Divergence (for $\bV$).
Each differential operator is written in spherical coordinates on the regular $(R, \theta)$ grid and being linear, is represented as a sparse matrix -- simplifying the algebraic manipulations.

@@

After substitution of boundary conditions, each differential equation is transformed into a sparse linear system of the form $\matr{A}\bx = \vect{b}$.
Due to coupling, it is not necessary to find the exact solution -- so we apply few relaxations (\ref{eq:Relax}) on each linear system, using an appropriate preconditioner $\matr{M}_\matr{A}^{-1}$:
\begin{eqnarray} \label{eq:Relax}
  \bx_{n+1} &=& \bx_{n} + \matr{M}_\matr{A}^{-1} \left(\vect{b} - \matr{A}\bx_{n}\right)
\end{eqnarray}

After the solver has converged to a solution for the coupled system, the total stress tensor $\mathbb{T}$ on the particle $\cl{S} = \left\{\bx : \|\bx\| = 1\right\}$ is computed (\ref{eq:Stress}) and integrated to yield the total force (\ref{eq:Force}):
\begin{eqnarray}
  \label{eq:Stress}
  \mathbb{T} &=& - P \mathbb{I} + \bnabla \bV + (\bnabla \bV) ^T  + \bnabla \varPhi \bnabla \varPhi - \frac{\bnabla \varPhi \cdot \bnabla \varPhi}{2} \mathbb{I}, \\
  \label{eq:Force}
  \vect{F} &=& \int_{\cl{S}} \left(\mathbb{T} \cdot \bnhat \right) ds = \boldsymbol{0}.
\end{eqnarray}
Since the problem is axisymmetric, the force should be aligned with the symmetry axis: $\vect{F} = F \bzhat$.
Moreover, since the particle is stress-free in steady flow, we have $F = 0$ for steady-state velocity $\cl{U}$.

This observation enables us to find the correct $\cl{U}$ for given $\beta$, using one-dimensional root-finding algorithm on $F_\beta(\cl{U})$, as a function of $\cl{U}$.

The numerical solution for $\cl{U}$ as a function of $\beta$ is compared to the closed-form solution from \cite{Yariv10b}.
Using a $60 \times 15$ grid for $(R,\theta) \in [1, 100] \times [0, \pi]$, we apply $2\mathord{,}000$ iterations in order to find the total force acting on the particle.
This step is repeated a few times to find the correct $\cl{U}$ so that $F_\beta(\cl{U}) = 0$ for various $\beta$ values -- see Figure \ref{fig:Results} for numerical results.
\begin{figure}[htbp]
\begin{framed}
    \begin{center}
        \includegraphics[width=0.45\textwidth]
            {convergence.eps}
        \includegraphics[width=0.45\textwidth]
            {comparison.eps}
        \caption{Left: Residual $\ell_2$ norm graph. Right: Particle Velocity $\cl{U}$ as a the function of electric field $\beta$.
        The error between the theoretical and the solver's results is less than 2\%.}
        \label{fig:Results}
    \end{center}
\end{framed}
\end{figure}

@@

$C$, $\varPhi$ and $P$ are discretized using a central grid whereas a staggered grid is used for $\bV$ at the problem domain $\cl{D}$ (see Figure \ref{fig:Grids}).
\begin{figure}[htbp]
\begin{framed}
    \begin{center}
        \includegraphics[width=1\textwidth]
            {StaggeredGrid.eps}
        \caption{Grids for $C$, $\varPhi$, $P$ and $\bV$.}
        \label{fig:Grids}
    \end{center}
\end{framed}
\end{figure}

Finite volume method is used to represent the partial differential equations in the form of an algebraic system, using central differences or an upwind scheme (for better stability).

``Ghost points'' are used to define Dirichlet and Neumann boundary conditions on
the problem domain's boundary $\partial \cl{D}$.

After boundary counditions are eliminated, Jacobi and Red-Black Gauss-Seidel preconditioners are used for equations \ref{eq:Laplace} and \ref{eq:Nernst}, where a Vanka-type \cite{Vanka86} preconditioner is used for equation \ref{eq:Stokes}.
The preconditioners are computed as sparse matrices, making the relaxations (at equation \ref{eq:Relax}) much more straightforward.

Vector extrapolation methods \cite{Sidi90} have been implemented and are used to accelerate the solver's convergence.
Multigrid solver \cite{Brandt77, Yavneh06} implementation will also be a part of the research.

>>>>

\title{M.Sc. Proposal}
\author{Roman Zeyde \\ \\  {\small Advisor: Prof. Irad Yavneh} }
\begin{document}
\maketitle
\section{Introduction}
Electrokinetic theory describes the dynamics of charged particles
in ionic fluids. When a particle acquires surface charge, a layer
of ions of opposite charge is attracted to the surface via
electric forces, creating a double-layer structure around the
particle (see Figure \ref{fig:EDL}). This structure, called
``Debye layer'', electrically screens the surface charge, thus
creating a potential difference between the particle and the outer
layer of the fluid bulk.

\begin{figure}[htbp]
\begin{framed}
    \begin{center}
        \includegraphics[width=0.25\textwidth]
            {ElectricDoubleLayer.eps}
        \caption{Schematic structure of the double layer:
        (1) particle surface, (2) Debye layer and (3) fluid bulk.
        If the particle surface has positive (blue) surface charge,
        it attracts negative (red) ions from the fluid making the
        Debye layer negatively charged (as opposed to the rest of
        the fluid bulk, which is electrically neutral).
        Zeta potential is defined as the voltage drop on (2).}
        \label{fig:EDL}
    \end{center}
\end{framed}
\end{figure}
In cases where the layer width is much smaller than the particle's
size, an analytical asymptotic solution for the Debye layer
dynamics can be found \cite{Yariv10a}.

Electrokinetics is described by the following physical phenomena (in dimensionless formulation):
\begin{enumerate}
\item Electrostatic potential---the fluid bulk contains no free
charge:
\begin{equation} \label{eq:Laplace}
    \bnabla \cdot (C \bnabla \varPhi) = 0 \, .
\end{equation}
\item Fluid dynamics---Stokes flow with an electrostatic force
component:
\begin{equation} \label{eq:Stokes}
\left\{ \begin{array}{l}
\bLaplacian \bV - \bnabla P + \Laplacian \varPhi \bnabla \varPhi = 0 \, , \\
\bnabla \cdot \bV = 0 \, . \end{array} \right.
\end{equation}
\item Ion diffusion and advection (via the fluid's velocity field):
\begin{equation} \label{eq:Nernst}
\Laplacian C - \alpha \bV \cdot \bnabla C = 0 \, .
\end{equation}
\end{enumerate}
{\bf IY: WRITE EXPLICITLY HERE WHAT THE VARIABLES ARE---C,P,V,
ETC.}The boundary conditions are determined by the specific
problem under consideration and defined by the particle's
geometry, chemical characteristics, and the fluid dynamics.

The partial differential equations above are coupled and
non-linear. In general, they have no analytic solution. Moreover,
any numerical solver must handle the scale disparity caused by the
Debye layer width being much smaller than the particle's size. A
closed-form linear asymptotic solution has been developed for
spherical ion-exchanging particles and weak electric field in an
axisymmetric setting \cite{Yariv10b} but it is hard to extend this
analytic solution to more general systems. Once the electric field
becomes stronger, significant nonlinear phenomena are expected.
This interesting regime has not yet been explored.

{\bf IY: ADD HERE A PARAGRAPH THAT DISCUSSES THE IMPORTANCE AND
RELEVANCE OF THE PROBLEM---YOU CAN TAKE THIS FROM THE PAPERS YOU
HAVE READ.}

\section{Research goals}
Our goal is to implement an accurate and fast iterative numerical
solver for electrokinetic problems, and then apply it to the study
of systems that have no closed-form solutions, including strong
electric fields, non-spherical particles, and asymmetric charge on
the particle surface. We expect to thus gain insight into the
chemical and physical behavior in far more general regimes than
are currently well-understood. We hope that this will also lead to
further analytical developments, based on new scaling behavior
provided by the numerical solver. We may also target interesting
phenomena that have been observed experimentally, e.g.,
electrophoretic\footnote{Electrophoresis is the physical
phenomenon of the motion of particles in an ion-containing fluid
under the influence of an electric field.} autonomous
micro-swimmers \cite{Paxton04, Howse07}.

The solver will be verified against known closed-form solutions
for each of the sub-problems listed above, as well as the full
coupled problem, analyzing the discrepancies between the
theoretical (asymptotic) and numerical results. Moreover, the
convergence rate of the solver will be optimized using numerical
acceleration methods in order to achieve high accuracy in
reasonable execution time.

\section{Preliminary results}
The numerical solver is being implemented as a MATLAB package,
which provides the needed tools for numerical solution of
electrokinetic problems. The solver applies an iterative
relaxation technique to solve the discretized coupled system.
Beginning with some initial guess for the solution The equations
are relaxed in order with respect to their corresponding
variables, with the remaining variables fixed at their current
values (``Gauss-Seidel style''):
\begin{itemize}
  \item The diffusion equation (\ref{eq:Laplace}) is used to update $\varPhi$.
  \item The Stokes system (\ref{eq:Stokes}) is used to update $\bV$ and $P$.
  \item The Nernst-Planck equation (\ref{eq:Nernst}) is used to update $C$.
\end{itemize}

The problem configuration is taken from \cite{Yariv10b} and
consists of a spherical ion-exchanging particle, surrounded by
infinite Stokes fluid. Spherical coordinates $(R,\theta,\phi)$ are
employed. The problem is axisymmetric, hence independent of
$\phi$, which simplifies the solution process. The electrokinetic
problem is thus written as a two-dimensional partial differential
equation system for $R \in [1,\infty)$ and $\theta \in [0, \pi]$.
Natural grid choices are a logarithmic grid for $R$ and a uniform
grid for $\theta$.

The following linear operators are implemented in spherical
coordinates on the grid described above: Scalar Laplacian (for
$\varPhi$ and $C$), Vector Laplacian (for $\bV$), Gradient (for
$\Phi, C$ and $P$) and Divergence (for $\bV$). Each differential
operator is written in spherical coordinates on the regular $(R,
\theta)$ grid and, being linear, is represented as a sparse
matrix, which is very convenient for the required algebraic
manipulations.

@@

Each differential equation and its boundary conditions are
discretized, resulting in a sparse linear system of the form
$\matr{A}\bx = \vect{b}$. We then cycle by applying a few
relaxation iterations on each linear system in turn, as noted
above, using an appropriate preconditioner
$\matr{M}_\matr{A}^{-1}$:
\begin{eqnarray} \label{eq:Relax}
  \bx_{n+1} &=& \bx_{n} + \matr{M}_\matr{A}^{-1} \left(\vect{b} - \matr{A}\bx_{n}\right)
\end{eqnarray}

After the iterations have converged to a solution for the coupled
system, the total stress tensor $\mathbb{T}$ on the particle
$\cl{S} = \left\{\bx : \|\bx\| = 1\right\}$ is computed
(\ref{eq:Stress}) and integrated to yield the total force
(\ref{eq:Force}):
\begin{eqnarray}
  \label{eq:Stress}
  \mathbb{T} &=& - P \mathbb{I} + \bnabla \bV + (\bnabla \bV) ^T  +
  \bnabla \varPhi \bnabla \varPhi - \frac{\bnabla \varPhi \cdot \bnabla \varPhi}{2} \mathbb{I}, \\
  \label{eq:Force}
  \vect{F} &=& \int_{\cl{S}} \left(\mathbb{T} \cdot \bnhat \right) ds = \boldsymbol{0}.
\end{eqnarray}
Because the problem is axisymmetric, the force should be aligned
with the symmetry axis: $\vect{F} = F \bzhat$. Moreover, since the
particle is stress-free in steady flow, we have $F = 0$ for
steady-state velocity $\cl{U}$. This observation enables us to
find the correct $\cl{U}$ for given $\beta$, using a
one-dimensional root-finding algorithm on $F_\beta(\cl{U})$, as a
function of $\cl{U}$.

The numerical solution for $\cl{U}$ as a function of $\beta$ is
compared to the closed-form asymptotic solution from
\cite{Yariv10b}. Using a $60 \times 15$ grid for $(R,\theta) \in
[1, 100] \times [0, \pi]$, we apply $2\mathord{,}000$ iterations
in order to find the total force acting on the particle. This step
is repeated a few times to find the correct $\cl{U}$ so that
$F_\beta(\cl{U}) = 0$ for various $\beta$ values---see Figure
\ref{fig:Results} for numerical results.
\begin{figure}[htbp]
\begin{framed}
    \begin{center}
        \includegraphics[width=0.45\textwidth]
            {convergence.eps}
        \includegraphics[width=0.45\textwidth]
            {comparison.eps}
        \caption{Left: Residual $\ell_2$ norm convergence graph. Right: Particle Velocity
        $\cl{U}$ as a the function of electric field $\beta$.
        The error between the theoretical and the solver's results is less than 2\%.}
        \label{fig:Results}
    \end{center}
\end{framed}
\end{figure}

@@

$C$, $\varPhi$ and $P$ are discretized at cell centers, whereas a
staggered grid is used for $\bV$ at the problem domain $\cl{D}$
(see Figure \ref{fig:Grids}).
\begin{figure}[htbp]
\begin{framed}
    \begin{center}
        \includegraphics[width=1\textwidth]
            {StaggeredGrid.eps}
        \caption{Grids for $C$, $\varPhi$, $P$ and $\bV$.}
        \label{fig:Grids}
    \end{center}
\end{framed}
\end{figure}
A finite volume discretization is used to represent the partial
differential equations in the form of an algebraic system, using
central differences or an upwind scheme (for stability). ``Ghost
points'' are used to define Dirichlet and Neumann boundary
conditions on the domain boundary $\partial \cl{D}$.

After boundary conditions are eliminated, Jacobi and Red-Black
Gauss-Seidel preconditioners are used for equations
(\ref{eq:Laplace}) and (\ref{eq:Nernst}), while a so-called
Vanka-type \cite{Vanka86} preconditioner is used for equation
(\ref{eq:Stokes}). The preconditioners are computed as sparse
matrices, facilitating the relaxation process (\ref{eq:Relax}).

Vector extrapolation methods \cite{Sidi90} have been implemented
and are used to accelerate the solver's convergence. Later, a
Multigrid framework \cite{Brandt77, Yavneh06} will be introduced
to allow the efficient solution of large problems.
